---
title: "Which Countries Are Most Vulnerable to AI-Driven Economic Disruption?"
subtitle: "A Global Comparative Analysis — Midterm Project, Advanced Modelling"
author: "Abdullah Tadmuri"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    theme: flatly
    highlight: zenburn
    code_folding: show
    fig_width: 10
    fig_height: 7
    css: style.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", dpi = 120
)
```

<div class="project-meta">
  <span class="meta-item"><strong>Scope:</strong> Global — all World Bank member countries</span>
  <span class="meta-sep">|</span>
  <span class="meta-item"><strong>Method:</strong> PCA + K-Means · Hierarchical · GMM Clustering</span>
  <span class="meta-sep">|</span>
  <span class="meta-item"><strong>Panel:</strong> 2015–2023</span>
  <span class="meta-sep">|</span>
  <span class="meta-item"><strong>Course:</strong> Advanced Modelling · UC3M · 2025–2026</span>
  <span class="meta-sep">|</span>
  <span class="meta-item"><strong>Source:</strong> World Bank WDI + Oxford AI Readiness Index</span>
</div>

# 1. Introduction

Artificial intelligence is no longer a distant prospect — it is actively reshaping industries, displacing routine tasks, and creating entirely new categories of work. But this transformation is not hitting every country equally. A software engineer in Stockholm and a subsistence farmer in sub-Saharan Africa face profoundly different futures, and the structural characteristics of their national economies — investment in research, educational attainment, digital infrastructure, labour market composition — will largely determine how smoothly each country navigates the transition.

This study asks a deceptively simple question with profound policy implications: *Which countries worldwide are most exposed to AI-driven economic disruption, and what makes some nations structurally more resilient than others?*

## 1.1 Research Question

**Which countries are most vulnerable to AI-driven economic disruption, and does vulnerability cluster systematically along the global development divide — separating high-income, middle-income, and low-income economies?**

To answer this, we take a data-driven, unsupervised learning approach. Rather than imposing a pre-defined vulnerability framework, we let the data reveal the underlying structure of the global economy. We use **Principal Component Analysis (PCA)** to reduce a set of correlated structural indicators (11 variables) into a smaller set of interpretable dimensions, and then apply **three independent clustering methods** — K-Means, Hierarchical, and Model-Based (Gaussian Mixture) — to group countries with similar structural profiles. The convergence (or divergence) of these methods provides a robust, multi-validated picture of the global economic landscape in the age of AI.

The analysis targets **all World Bank member countries** for which sufficient data is available across the 2015–2023 panel. Countries are stratified by the World Bank's income classification into three development tiers: **High-Income (analogous to the "1st World")**, **Middle-Income (Upper and Lower, analogous to the "2nd World")**, and **Low-Income (analogous to the "3rd World")**. This tripartite division allows us to ask whether the vulnerability patterns identified by our unsupervised methods mirror, or diverge from, the conventional development hierarchy — and to generate policy recommendations that are sensitive to the very different capacities of rich and poor nations to adapt to technological disruption.

## 1.2 Data Sources

Our analysis draws on three complementary data sources, each contributing different dimensions of the vulnerability picture:

| Source | Variables | Type |
|-----------------------|-------------------------------|------------------|
| World Bank WDI | R&D intensity, high-tech exports, tertiary education, GDP per capita, GDP growth, labour force participation, internet users, unemployment (total & youth), agricultural employment, industrial employment | Panel (2015–2023) |
| Kaggle | AI automation rate, AI readiness score (synthetic) | Panel (2015–2025) |
| Oxford Insights | Government AI Readiness Index 2025 (6 pillars) | Cross-section (2025) |

The **World Bank World Development Indicators (WDI)** database is our exclusive primary source, providing internationally standardised macroeconomic, human capital, digital infrastructure, labour market, and employment structure variables for virtually every country in the world. Sourcing all structural indicators from a single, unified database eliminates the inconsistencies and coverage gaps that arise from combining multiple national statistical agencies. The **Kaggle** dataset supplements these with AI-specific synthetic metrics (treated as illustrative only). The **Oxford Insights** Government AI Readiness Index adds a policy dimension and, crucially, is itself a global index covering approximately 160 countries — making it more useful here than in a purely European analysis.

Sources: World Bank DataBank (databank.worldbank.org); Oxford Insights (2025) *Government AI Readiness Index*; Kaggle *Global AI Workforce Automation 2015–2025* dataset.

------------------------------------------------------------------------

# 2. Setup & Data Collection

This section handles the technical groundwork: loading R packages, defining helper functions, and pulling raw data from the World Bank API. The entire data collection pipeline is fully reproducible — running this code from a fresh R session will automatically fetch the latest available data and produce a unified panel ready for analysis.

We load packages spanning five categories: **data wrangling** (tidyverse, janitor, countrycode), **API access** (WDI for World Bank), **dimensionality reduction and clustering** (FactoMineR, factoextra, NbClust, cluster, mclust), **visualisation** (corrplot, dendextend, ggrepel), and **geospatial mapping** (sf, rnaturalearth).

```{r packages}
pkgs <- c(
  "tidyverse", "janitor", "WDI", "countrycode",
  "FactoMineR", "factoextra", "NbClust", "cluster", "mclust",
  "psych", "dendextend", "corrplot",
  "sf", "rnaturalearth", "rnaturalearthdata", "ggrepel",
  "knitr", "kableExtra"
)
install.packages(setdiff(pkgs, installed.packages()[, "Package"]), quiet = TRUE)
invisible(lapply(pkgs, library, character.only = TRUE))
```

```{r helpers}
# Standardize to z-scores (used throughout)
z <- function(x) as.numeric(scale(x))

# Map World Bank income group to a three-tier development classification
income_to_tier <- function(income_group) {
  case_when(
    income_group == "High income"                                          ~ "High Income (1st World)",
    income_group %in% c("Upper middle income", "Lower middle income")     ~ "Middle Income (2nd World)",
    income_group == "Low income"                                           ~ "Low Income (3rd World)",
    TRUE                                                                   ~ NA_character_
  )
}

START <- 2015L
END   <- 2023L
```

## 2.1 World Bank WDI — Structural Indicators (All Countries)

The World Bank's World Development Indicators (WDI) database is our sole primary source of structural data. By sourcing all indicators from a single database, we avoid the cross-source inconsistencies that arise when merging national statistical agencies with different methodologies, coverage periods, and country definitions.

We retrieve eleven indicators that collectively capture a country's capacity to generate, absorb, and benefit from technological innovation, alongside its exposure to automation-driven disruption:

- **R&D Intensity** (% of GDP): Measures national commitment to research and development — a direct proxy for innovation capacity and the ability to create AI rather than merely be displaced by it.
- **High-Tech Exports** (% of manufactured exports): Reflects the economy's current positioning in knowledge-intensive trade — a signal of structural sophistication.
- **Tertiary Education** (% of population 25+ with at least a bachelor's degree): Captures the human capital base available to adapt to AI-driven changes.
- **GDP per Capita** (constant 2015 USD): A broad measure of economic development and the fiscal capacity to invest in adjustment policies.
- **GDP Growth** (%): Captures economic dynamism and recent trajectory.
- **Labour Force Participation** (%): Measures how much of the working-age population is economically active.
- **Internet Users** (% of population): A direct measure of digital infrastructure penetration at the population level — countries where the internet is widespread have a larger pool of digitally literate workers.
- **Unemployment Rate** (% of total labour force, ILO modelled estimate): A key labour market health indicator.
- **Youth Unemployment Rate** (% of labour force ages 15–24): High youth unemployment signals labour market rigidity and weak absorptive capacity — both of which amplify disruption risk.
- **Agricultural Employment** (% of total employment): The share of the workforce in agriculture — a sector characterised by low digitalisation, high manual-task intensity, and significant automation exposure.
- **Industrial Employment** (% of total employment): The share in industry (mining, manufacturing, construction, utilities) — a sector with large and growing automation exposure, particularly in assembly-line and repetitive manufacturing tasks.

The `extra = TRUE` flag retrieves additional country metadata including the World Bank income group and geographic region, which we use to define our development tier classification.

```{r fetch-wdi, cache=TRUE}
wdi_raw <- WDI(
  indicator = c(
    RD_Intensity   = "GB.XPD.RSDV.GD.ZS",
    HiTech_Exports = "TX.VAL.TECH.MF.ZS",
    Tertiary_Edu   = "SE.TER.CUAT.BA.ZS",
    GDP_pc         = "NY.GDP.PCAP.KD",
    GDP_Growth     = "NY.GDP.MKTP.KD.ZG",
    Labor_Part     = "SL.TLF.CACT.ZS",
    Internet_Users = "IT.NET.USER.ZS",
    Unemp_Total    = "SL.UEM.TOTL.ZS",
    Unemp_Youth    = "SL.UEM.1524.ZS",
    Agric_Empl     = "SL.AGR.EMPL.ZS",
    Industry_Empl  = "SL.IND.EMPL.ZS"
  ),
  country = "all",
  start = START, end = END,
  extra = TRUE
) |>
  # Remove regional and income-group aggregates; keep country-level observations only
  filter(
    !is.na(income),
    income != "Aggregates",
    str_length(iso3c) == 3
  ) |>
  mutate(
    Development_Tier = income_to_tier(income),
    Development_Tier = factor(
      Development_Tier,
      levels = c("High Income (1st World)", "Middle Income (2nd World)", "Low Income (3rd World)")
    )
  ) |>
  select(iso3c, country, year, region, income, Development_Tier,
         RD_Intensity, HiTech_Exports, Tertiary_Edu,
         GDP_pc, GDP_Growth, Labor_Part,
         Internet_Users, Unemp_Total, Unemp_Youth,
         Agric_Empl, Industry_Empl)

cat("WDI panel:", nrow(wdi_raw), "rows,",
    n_distinct(wdi_raw$iso3c), "countries,",
    min(wdi_raw$year), "-", max(wdi_raw$year), "\n")
cat("Development Tier breakdown:\n")
wdi_raw |>
  filter(year == END) |>
  count(Development_Tier) |>
  print()
```

## 2.2 Country Classification & Scope

The World Bank income group — retrieved above via `extra = TRUE` — provides our primary country stratification. We define three development tiers that map onto the conventional "worlds" terminology while using the World Bank's annually updated, income-threshold-based definitions:

| Development Tier | World Bank Income Group | Approx. GNI per Capita (2023) | Interpretation |
|-----------------|------------------------|-------------------------------|----------------|
| High Income (1st World) | High income | > $13,845 | Advanced economies: OECD members, wealthy Gulf states, Singapore, etc. |
| Middle Income (2nd World) | Upper + Lower middle | $1,136–$13,845 | Emerging economies: China, Brazil, India, Indonesia, South Africa, etc. |
| Low Income (3rd World) | Low income | < $1,136 | Least-developed economies: Sub-Saharan Africa, parts of South/Southeast Asia |

This classification is operationally grounded — unlike the original Cold War "worlds" terminology, which was a geopolitical rather than economic concept, the World Bank income groups are recalculated annually using GNI per capita thresholds (Atlas method). They are the standard international reference for development-level comparisons.

## 2.3 Supplementary: Kaggle & Oxford Insights

We supplement the official World Bank statistics with two additional sources. The **Kaggle AI Workforce Automation** dataset provides country-level estimates of automation rates and AI readiness scores. It is important to note that this dataset is *synthetically generated* — the values are modelled rather than observed — so we treat it strictly as an illustrative supplement. The **Oxford Insights Government AI Readiness Index** adds a policy dimension, scoring approximately 160 governments on their preparedness for AI across six pillars (governance, infrastructure, data availability, human capital, innovation, and government capacity). Unlike the European analysis it was previously used in, here its global coverage is a genuine asset.

**Note on data files:** The supplementary CSV files should be placed in a `data/` subfolder within the same directory as this Rmd file. The zip submission should include both CSVs.

```{r supplementary}
# Kaggle AI automation panel (synthetic — for external validation only)
ai_auto <- tryCatch({
  read_csv(file.path("data", "global_ai_workforce_automation_2015_2025.csv"),
           show_col_types = FALSE) |>
    clean_names() |>
    transmute(
      iso3c = countrycode(country, "country.name", "iso3c", warn = FALSE),
      year  = as.integer(year),
      automation_rate    = automation_rate_percent,
      ai_readiness_score = ai_readiness_score
    ) |>
    filter(!is.na(iso3c))
}, error = function(e) {
  message("Kaggle file not found in data/ — external validation will be skipped. ", e$message)
  tibble(iso3c = character(), year = integer(),
         automation_rate = numeric(), ai_readiness_score = numeric())
})

# Oxford Insights Government AI Readiness Index 2025 (global cross-section)
ai_readiness <- tryCatch({
  raw_ox <- read_csv(file.path("data", "ai_readiness_index.csv"),
                     show_col_types = FALSE) |>
    clean_names()   # normalises column names to snake_case (e.g., "Overall Score" -> "overall_score")

  # Robustly resolve ISO3 codes whether the file has them or only country names
  if ("iso3c" %in% names(raw_ox)) {
    raw_ox <- raw_ox |> mutate(iso3c = as.character(iso3c))
  } else {
    raw_ox <- raw_ox |>
      mutate(iso3c = countrycode(country, "country.name", "iso3c", warn = FALSE))
  }

  raw_ox |>
    select(iso3c, AI_Readiness = overall_score) |>
    filter(!is.na(iso3c), !is.na(AI_Readiness))
}, error = function(e) {
  message("Oxford AI Readiness file not found in data/ — external validation will be skipped. ", e$message)
  tibble(iso3c = character(), AI_Readiness = numeric())
})

cat("Kaggle records loaded:", nrow(ai_auto), "\n")
cat("Oxford AI Readiness records loaded:", nrow(ai_readiness), "\n")
```

## 2.4 Merge Master Panel

With the primary and supplementary data loaded, we merge them into a single master panel dataset — one row per country per year. The join is performed on ISO-3 country codes and year. Supplementary datasets are joined with `left_join()` so that their absence does not reduce the core panel. Where multiple records exist for the same country-year, we take the mean.

```{r merge}
all_data <- wdi_raw |>
  left_join(ai_auto,      by = c("iso3c", "year")) |>
  left_join(ai_readiness, by = "iso3c") |>
  group_by(iso3c, country, year, region, income, Development_Tier) |>
  summarise(across(where(is.numeric), \(x) mean(x, na.rm = TRUE)),
            .groups = "drop")

cat("Master panel:", nrow(all_data), "rows,",
    n_distinct(all_data$iso3c), "countries,",
    min(all_data$year), "-", max(all_data$year), "\n")
```

------------------------------------------------------------------------

# 3. Pre-Processing & Descriptive Analysis

Before applying any statistical technique, we must thoroughly understand the structure, quality, and distributional properties of our data. This section serves as a diagnostic checkpoint: we examine the data dimensions, assess missing values, define our analysis variables, inspect distributions, detect outliers, and explore the correlation structure. Each step informs critical decisions downstream.

## 3.1 Data Structure

A quick structural overview confirms the panel format and verifies that all expected variables are present.

```{r structure}
glimpse(all_data)
```

## 3.2 Missing Data Analysis

Missing data is an unavoidable reality when working with global data. High-income countries tend to report comprehensively to the World Bank, while low-income countries — where reporting infrastructure is weaker — often have gaps in R&D spending, tertiary education attainment, and high-technology trade statistics. Understanding the pattern of missingness is critical: if gaps are systematic by development tier (as is likely), complete-case analysis may underrepresent low-income countries.

```{r missing-data, fig.height=6}
analysis_core <- c("RD_Intensity", "HiTech_Exports", "Tertiary_Edu",
                   "GDP_pc", "GDP_Growth", "Labor_Part",
                   "Internet_Users", "Unemp_Total", "Unemp_Youth",
                   "Agric_Empl", "Industry_Empl")

miss_pct <- all_data |>
  select(all_of(analysis_core)) |>
  summarise(across(everything(), \(x) mean(is.na(x)) * 100)) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Pct_Missing") |>
  arrange(desc(Pct_Missing))

ggplot(miss_pct, aes(reorder(Variable, Pct_Missing), Pct_Missing)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = sprintf("%.1f%%", Pct_Missing)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  labs(title = "Missing Data by Variable (Full Panel, All Countries)",
       x = NULL, y = "% Missing") +
  theme_minimal() +
  ylim(0, max(miss_pct$Pct_Missing) * 1.2)
```

```{r missing-by-tier, fig.height=5}
# Missingness by development tier highlights the systematic data gap
all_data |>
  filter(!is.na(Development_Tier)) |>
  select(Development_Tier, all_of(analysis_core)) |>
  group_by(Development_Tier) |>
  summarise(across(everything(), \(x) mean(is.na(x)) * 100), .groups = "drop") |>
  pivot_longer(-Development_Tier, names_to = "Variable", values_to = "Pct_Missing") |>
  ggplot(aes(Variable, Pct_Missing, fill = Development_Tier)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#2196F3", "#FF9800", "#F44336")) +
  labs(title = "Missing Data by Variable and Development Tier",
       subtitle = "Low-income countries have systematically higher missingness for R&D and education variables",
       x = NULL, y = "% Missing", fill = "Tier") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r missing-table}
miss_pct |>
  kable(digits = 1, col.names = c("Variable", "% Missing"),
        caption = "Overall Missing Data by Variable") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## 3.3 Feature Engineering & Complete-Case Filter

The eleven WDI variables used here are expressed as ratios or percentages of well-defined totals (ILO modelled estimates, total employment, total population), making them directly comparable across countries of all sizes. Germany and Luxembourg are compared on equal footing because all variables capture *structure*, not scale. This is a key advantage of the global WDI approach.

We define our final analysis variable set and apply a **complete-case filter**: only country-year observations with valid values on all eleven analysis variables are retained. While this reduces sample size — particularly for low-income countries where R&D and tertiary education reporting is sparse — it ensures that PCA operates on a consistent, fully-observed matrix, which is a mathematical requirement for accurate eigenvalue decomposition.

```{r feature-engineering}
analysis_vars <- c(
  "RD_Intensity",   # Innovation capacity
  "HiTech_Exports", # Knowledge-economy positioning
  "Tertiary_Edu",   # Human capital
  "GDP_pc",         # Economic development level
  "GDP_Growth",     # Economic dynamism (cyclical)
  "Labor_Part",     # Labour market activity
  "Internet_Users", # Digital infrastructure penetration
  "Unemp_Total",    # Labour market distress (structural)
  "Unemp_Youth",    # Youth labour market vulnerability
  "Agric_Empl",     # Agricultural employment share (ILO % of total)
  "Industry_Empl"   # Industrial employment share (ILO % of total)
)

df_cc <- all_data |>
  filter(year >= START, year <= END,
         !is.na(Development_Tier)) |>
  filter(!if_any(all_of(analysis_vars), is.na))

cat("Complete-case panel:", nrow(df_cc), "rows,",
    n_distinct(df_cc$iso3c), "countries,",
    length(analysis_vars), "analysis variables\n")
cat("Variables:", paste(analysis_vars, collapse = ", "), "\n")
```

**Result:** The final analysis dataset contains `r nrow(df_cc)` observations (country-years), `r n_distinct(df_cc$iso3c)` countries, and `r length(analysis_vars)` structural indicators — satisfying the project requirements of $n > 100$ observations and $p \geq 10$ variables.

```{r stats-banner, echo=FALSE, results='asis'}
cat(sprintf('
<div class="stat-grid">
  <div class="stat-card">
    <div class="stat-value">%d</div>
    <div class="stat-label">Countries</div>
  </div>
  <div class="stat-card">
    <div class="stat-value">%d</div>
    <div class="stat-label">Variables</div>
  </div>
  <div class="stat-card">
    <div class="stat-value">%d</div>
    <div class="stat-label">Years (2015–2023)</div>
  </div>
  <div class="stat-card">
    <div class="stat-value">3</div>
    <div class="stat-label">Clustering Methods</div>
  </div>
  <div class="stat-card">
    <div class="stat-value">%s</div>
    <div class="stat-label">Observations</div>
  </div>
</div>
',
n_distinct(df_cc$iso3c),
length(analysis_vars),
as.integer(END - START + 1),
formatC(nrow(df_cc), format = "d", big.mark = ",")
))
```

**A note on the Kaggle and Oxford Insights data:** These datasets were deliberately excluded from the core PCA/clustering pipeline. The Kaggle data is *synthetically generated*, so incorporating it alongside official statistics would contaminate the factor structure with artificial patterns. The Oxford Insights index is *cross-sectional* (one score per country, 2025 only), which is incompatible with our panel-based PCA. Both are used later as **external validators** in Section 5.8.

### Countries Included and Excluded

```{r country-coverage}
# all_data already carries Development_Tier from the WDI fetch — use it directly
all_targeted <- all_data |>
  distinct(iso3c, country, Development_Tier) |>
  arrange(country)

included <- df_cc |>
  distinct(iso3c, country, Development_Tier) |>
  arrange(country)

excluded <- all_data |>
  distinct(iso3c, country, Development_Tier) |>
  filter(!iso3c %in% included$iso3c) |>
  arrange(Development_Tier, country)

cat("Countries in source data:", n_distinct(all_data$iso3c), "\n")
cat("Countries in final analysis:", nrow(included), "\n")
cat("Countries excluded:", nrow(excluded), "\n\n")

excluded |>
  count(Development_Tier, name = "n_excluded") |>
  kable(caption = "Excluded Countries by Development Tier") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

The table above confirms the anticipated pattern: the majority of excluded countries are in the Low-Income tier, primarily because R&D spending and tertiary education attainment data are not systematically reported to the World Bank by the world's least-developed economies. This systematic attrition is acknowledged as a limitation in Section 8.2 — the vulnerability rankings will under-represent the very poorest countries, which may in fact be the most vulnerable.

### Development Tier Coverage in Final Analysis

```{r tier-coverage}
df_cc |>
  filter(year == END) |>
  count(Development_Tier, name = "n_countries") |>
  kable(caption = paste("Countries in Final Analysis by Development Tier (", END, ")")) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## 3.4 Summary Statistics

The descriptive statistics below provide a concise portrait of our analysis variables. For each indicator, we report the mean, standard deviation, quartiles, and range.

```{r summary-stats}
df_cc |>
  select(all_of(analysis_vars)) |>
  pivot_longer(everything(), names_to = "Variable") |>
  group_by(Variable) |>
  summarise(
    Mean   = mean(value),
    SD     = sd(value),
    Min    = min(value),
    Q1     = quantile(value, 0.25),
    Median = median(value),
    Q3     = quantile(value, 0.75),
    Max    = max(value),
    .groups = "drop"
  ) |>
  kable(digits = 2, caption = "Summary Statistics of Analysis Variables") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Mean Values by Development Tier

This pre-PCA comparison already reveals the broad structural differences between development levels that PCA will subsequently formalise.

```{r tier-means, fig.height=9}
df_cc |>
  select(Development_Tier, all_of(analysis_vars)) |>
  pivot_longer(-Development_Tier, names_to = "Variable", values_to = "Value") |>
  group_by(Development_Tier, Variable) |>
  summarise(Mean = mean(Value), .groups = "drop") |>
  ggplot(aes(Development_Tier, Mean, fill = Development_Tier)) +
  geom_col(alpha = 0.85) +
  scale_fill_manual(values = c("#2196F3", "#FF9800", "#F44336")) +
  facet_wrap(~ Variable, scales = "free_y", ncol = 3) +
  labs(title = "Mean Variable Values by Development Tier",
       subtitle = "Pre-PCA comparison: structural divergence across the global development hierarchy",
       x = NULL, y = "Mean Value", fill = "Tier") +
  theme_minimal() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())
```

## 3.5 Distributions

Examining the distributional shape of each variable is essential for PCA, which works best when variables are roughly normally distributed. Variables like GDP per capita typically show pronounced right skew in a global sample — a handful of very wealthy countries pull the distribution tail rightward. Standardising to z-scores (which PCA does automatically via `scale.unit = TRUE`) corrects for differences in scale but does not eliminate skewness.

```{r distributions, fig.height=11}
df_cc |>
  select(all_of(analysis_vars)) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") |>
  ggplot(aes(Value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30,
                 fill = "steelblue", alpha = 0.7) +
  geom_density(colour = "darkred", linewidth = 0.8) +
  facet_wrap(~ Variable, scales = "free", ncol = 3) +
  labs(title = "Distribution of Analysis Variables (All Country-Years)") +
  theme_minimal()
```

## 3.6 Outlier Detection

We use a z-score criterion (|z| > 3) to flag potential outliers. In a normal distribution, fewer than 0.3% of observations exceed this threshold, so any flagged values genuinely stand out. We retain all observations — extreme but real country-level values (e.g., Singapore's extraordinary high-tech export share, or a Gulf state's very low agricultural employment) represent genuine structural characteristics, not data errors.

```{r outliers}
z_outliers <- df_cc |>
  select(all_of(analysis_vars)) |>
  mutate(across(everything(), z)) |>
  summarise(across(everything(), \(x) sum(abs(x) > 3))) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Outliers_z3") |>
  arrange(desc(Outliers_z3))

z_outliers |>
  kable(caption = "Number of Extreme Observations (|z| > 3) per Variable") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## 3.7 Correlation Analysis

The correlation matrix is the single most important diagnostic for PCA suitability. PCA works by compressing correlated variables into fewer dimensions — if variables are uncorrelated, there is nothing to compress. Conversely, strong correlations among groups of variables indicate redundant information that PCA can efficiently consolidate.

The heatmap below displays Pearson correlation coefficients between all analysis variables.

```{r correlation, fig.height=8}
cor_mat <- cor(df_cc[, analysis_vars], use = "pairwise.complete.obs")

corrplot(cor_mat, method = "color", type = "upper",
         addCoef.col = "black", number.cex = 0.65,
         tl.col = "black", tl.srt = 45, tl.cex = 0.8,
         title = "Correlation Matrix of Analysis Variables", mar = c(0, 0, 2, 0))
```

**Key patterns from the global correlation structure:**

- **A "development" cluster**: GDP per capita correlates strongly with R&D intensity (r=`r round(cor_mat["GDP_pc","RD_Intensity"], 2)`), internet users (r=`r round(cor_mat["GDP_pc","Internet_Users"], 2)`), and tertiary education (r=`r round(cor_mat["GDP_pc","Tertiary_Edu"], 2)`). This mirrors the well-established co-movement of economic development indicators globally — wealthier countries simultaneously invest more in R&D, educate more of their population to degree level, and have more of their citizens online.
- **A "structural vulnerability" cluster**: Agricultural employment and total unemployment are positively correlated (r=`r round(cor_mat["Agric_Empl","Unemp_Total"], 2)`), and both are strongly negatively associated with GDP per capita. Countries with large agricultural sectors tend to be poorer, have weaker labour markets, and show lower digital infrastructure.
- **A "youth labour market" signal**: Youth unemployment correlates strongly with total unemployment (r=`r round(cor_mat["Unemp_Youth","Unemp_Total"], 2)`) and negatively with GDP per capita, suggesting that young people in low-income countries face especially difficult labour market entry.

These strong, interpretable correlation structures confirm that the data is well-suited for PCA — there is substantial shared variance to be compressed into a smaller number of meaningful dimensions.

------------------------------------------------------------------------

# 4. Principal Component Analysis (PCA)

Principal Component Analysis is the backbone of our dimensionality reduction strategy. With `r length(analysis_vars)` correlated structural indicators, direct analysis would be unwieldy and prone to multicollinearity issues. PCA transforms the original variables into a new set of **principal components** — uncorrelated linear combinations ordered by the amount of variance they capture. The first component explains the most variance; each subsequent component captures the most remaining variance orthogonal to all preceding components.

The practical benefit is twofold. First, PCA allows us to *visualise* the multi-dimensional structure of global economies on simple 2D plots. Second, clustering in PCA space reduces noise from minor components and avoids the "curse of dimensionality" — a phenomenon where distance-based methods degrade as the number of dimensions grows, because all pairs of points become nearly equidistant.

PCA is performed on **standardised (z-scored) data** (`scale.unit = TRUE`) so that variables on very different scales (e.g., GDP per capita in thousands of dollars vs. an employment share in percentages) contribute equally to the analysis.

## 4.1 Data Suitability

We apply two formal diagnostic tests before computing PCA:

- The **Kaiser-Meyer-Olkin (KMO)** measure of sampling adequacy quantifies the proportion of variance among variables that might be caused by underlying common factors. Values above 0.5 are acceptable, above 0.7 good, and above 0.8 excellent (Kaiser 1974).
- **Bartlett's test of sphericity** tests the null hypothesis that the correlation matrix is an identity matrix (all variables independent). A significant result ($p < 0.001$) confirms that meaningful correlations exist to exploit.

```{r kmo-bartlett}
kmo_result     <- KMO(cor_mat)
bartlett_result <- cortest.bartlett(cor_mat, n = nrow(df_cc))

cat("KMO overall MSA:", round(kmo_result$MSA, 3), "\n")
cat("Bartlett's test: chi-sq =", round(bartlett_result$chisq, 1),
    ", df =", bartlett_result$df,
    ", p =", format.pval(bartlett_result$p.value, digits = 3), "\n")
```

Both tests should confirm strong suitability: the globally correlated nature of development indicators (education, income, R&D, internet, employment structure) typically produces KMO values well above 0.8 in samples of this kind.

## 4.2 PCA Computation

```{r pca-compute}
# ncp set to the full number of variables to ensure no components are truncated
pca_res <- PCA(df_cc[, analysis_vars], scale.unit = TRUE,
               ncp = length(analysis_vars), graph = FALSE)

eig_tbl <- get_eigenvalue(pca_res) |>
  as.data.frame() |>
  rownames_to_column("PC") |>
  as_tibble()

eig_tbl |>
  kable(digits = 2,
        col.names = c("Component", "Eigenvalue", "% Variance", "Cumulative %"),
        caption = "PCA Eigenvalues — Full Decomposition") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## 4.3 Scree Plot

The scree plot visualises eigenvalues in descending order. We apply the **Kaiser criterion** (retain all components with eigenvalue $\geq 1$, shown as the red dashed line) as our primary retention rule: a component must capture more variance than any single original standardised variable to be worth retaining. The visual "elbow" in the plot provides a complementary guide.

```{r scree, fig.height=5}
ggplot(eig_tbl, aes(x = seq_len(nrow(eig_tbl)), y = eigenvalue)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "red", linewidth = 0.8) +
  geom_text(aes(label = paste0(round(variance.percent, 1), "%")),
            vjust = -0.5, size = 3.2) +
  scale_x_continuous(breaks = seq_len(nrow(eig_tbl)),
                     labels = paste0("PC", seq_len(nrow(eig_tbl)))) +
  labs(title = "Scree Plot with Kaiser Criterion (eigenvalue = 1)",
       x = "Principal Component", y = "Eigenvalue") +
  theme_minimal()

n_retain <- sum(eig_tbl$eigenvalue >= 1)
cum_var  <- eig_tbl$cumulative.variance.percent[n_retain]
cat("Components retained (Kaiser criterion):", n_retain,
    "| Cumulative variance:", round(cum_var, 1), "%\n")
```

By the **Kaiser criterion**, we retain **`r n_retain` components** explaining **`r round(cum_var, 1)`%** of total variance.

## 4.4 Loadings & Interpretation

**Loadings** are the correlations between original variables and each principal component. High positive loadings indicate the variable increases with the component; high negative loadings indicate an inverse relationship. We report `$coord` from FactoMineR, which equals the variable-PC correlations in a standardised PCA (`scale.unit = TRUE`), providing directly interpretable factor loadings.

```{r loadings}
load_df <- get_pca_var(pca_res)$coord |>
  as.data.frame() |>
  rownames_to_column("Variable") |>
  as_tibble()

# Display loadings for retained components only
load_df |>
  select(Variable, all_of(paste0("Dim.", 1:min(n_retain, 5)))) |>
  arrange(desc(abs(Dim.1))) |>
  kable(digits = 3,
        caption = "PCA Loadings: Variable–Component Correlations (retained components)") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

**Interpretation of retained components (threshold: |loading| > 0.4):**

```{r interpret-pca, results='asis'}
for (d in 1:min(n_retain, 3)) {
  col <- paste0("Dim.", d)
  top_pos <- load_df |> filter(.data[[col]] > 0.4) |> arrange(desc(.data[[col]]))
  top_neg <- load_df |> filter(.data[[col]] < -0.4) |> arrange(.data[[col]])

  cat(sprintf("\n- **PC%d** (%.1f%% variance): ",
              d, eig_tbl$variance.percent[d]))
  if (nrow(top_pos) > 0)
    cat("Positively loaded on ", paste(top_pos$Variable, collapse = ", "), ". ", sep = "")
  if (nrow(top_neg) > 0)
    cat("Negatively loaded on ", paste(top_neg$Variable, collapse = ", "), ". ", sep = "")
  cat("\n")
}
```

In global cross-country data, the dominant first component almost invariably captures the **global development axis** — a single underlying dimension that reflects the deep co-movement of income, education, R&D, digital infrastructure, and sectoral composition. Countries scoring high on PC1 tend to be wealthy, educated, R&D-intensive, digitally connected, and have small agricultural sectors; those scoring low show the opposite profile.

## 4.5 Variable Contributions

**Contributions** measure the magnitude of each variable's role in defining each component, expressed as a percentage. The red dashed reference line marks the expected contribution if all variables contributed equally (100% ÷ number of variables).

```{r var-contrib, fig.height=6}
fviz_contrib(pca_res, choice = "var", axes = 1, top = length(analysis_vars),
             title = "Variable Contributions to PC1") +
  theme_minimal()
```

```{r var-contrib-2, fig.height=6}
fviz_contrib(pca_res, choice = "var", axes = 2, top = length(analysis_vars),
             title = "Variable Contributions to PC2") +
  theme_minimal()
```

## 4.6 PCA Variable Map (cos²)

The variable map projects original variables onto the first two principal components. Variables appearing close together are positively correlated; those on opposite sides of the origin are negatively correlated; those at right angles are uncorrelated. The colour gradient represents $\cos^2$ — the quality of each variable's representation in this 2D plane.

```{r pca-var-map, fig.height=7}
fviz_pca_var(pca_res, col.var = "cos2", repel = TRUE,
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             title = "PCA Variable Map — coloured by cos² (quality of 2D representation)")
```

## 4.7 Biplot — Country Positions Coloured by Development Tier

This biplot projects each country's structural profile onto PC1–PC2. Points are coloured by World Bank income tier, allowing us to see immediately whether the data-driven PCA dimensions align with the conventional development hierarchy. If PCA has successfully captured the global development axis in PC1, we would expect a clear left-to-right gradient from Low-Income to High-Income countries.

Only the **latest available year** (`r max(df_cc$year)`) is shown to avoid overplotting.

```{r biplot, fig.height=8, fig.width=10}
ind_scores <- as_tibble(pca_res$ind$coord[, 1:min(n_retain, 5)])
names(ind_scores) <- paste0("PC", seq_along(ind_scores))
ind_scores$row_id <- seq_len(nrow(ind_scores))

df_pca <- df_cc |>
  mutate(row_id = row_number()) |>
  left_join(ind_scores, by = "row_id") |>
  select(-row_id)

latest_year <- max(df_pca$year)
df_latest   <- df_pca |> filter(year == latest_year)

ggplot(df_latest, aes(PC1, PC2, colour = Development_Tier, label = iso3c)) +
  geom_point(size = 3, alpha = 0.85) +
  geom_text_repel(size = 2.8, max.overlaps = 30) +
  scale_colour_manual(
    values = c("High Income (1st World)"   = "#2196F3",
               "Middle Income (2nd World)" = "#FF9800",
               "Low Income (3rd World)"    = "#F44336")
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", alpha = 0.5) +
  labs(
    title = paste("PCA Biplot — Country Positions (", latest_year, ")"),
    subtitle = "Coloured by World Bank Income Group",
    x = sprintf("PC1 (%.1f%%)", eig_tbl$variance.percent[1]),
    y = sprintf("PC2 (%.1f%%)", eig_tbl$variance.percent[2]),
    colour = "Development Tier"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## 4.8 Representation Quality (cos²)

The $\cos^2$ statistic measures how accurately each country's position in the 2D biplot represents its full multi-dimensional profile. Values below 0.5 (shown in red) indicate that important aspects of that country's economy are captured by higher-order components, and the 2D position may be misleading.

```{r cos2, fig.height=9}
ind_cos2 <- get_pca_ind(pca_res)$cos2 |>
  as.data.frame() |>
  mutate(cos2_12 = Dim.1 + Dim.2, row_id = row_number())

df_cos2 <- df_cc |>
  mutate(row_id = row_number()) |>
  left_join(ind_cos2 |> select(row_id, cos2_12), by = "row_id") |>
  filter(year == latest_year) |>
  arrange(cos2_12) |>
  mutate(bar_fill = if_else(cos2_12 > 0.5, "steelblue", "tomato"))

ggplot(df_cos2, aes(reorder(iso3c, cos2_12), cos2_12, fill = bar_fill)) +
  geom_col() +
  scale_fill_identity() +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "red") +
  coord_flip() +
  labs(title = "cos² Quality of Representation on PC1–PC2",
       subtitle = "Red bars: poorly represented in 2D (cos² < 0.5) — interpret biplot position with caution",
       x = NULL, y = "cos² (PC1 + PC2)") +
  theme_minimal()
```

------------------------------------------------------------------------

# 5. Clustering

With PCA complete, we now identify **groups** of countries that share similar structural profiles. While PCA revealed the main dimensions of global economic variation, clustering takes the next step: assigning countries to discrete, interpretable groups. We apply three fundamentally different clustering algorithms — **K-Means**, **Hierarchical (agglomerative)**, and **Model-Based (Gaussian Mixture)** — to exploit their complementary assumptions. Agreement across methods provides strong evidence that the groups are genuine features of the data.

All clustering is performed in **PCA space** (the first `r n_retain` retained components). This eliminates multicollinearity between correlated features, reduces noise from uninformative minor components, and mitigates the curse of dimensionality.

```{r cluster-data}
n_pcs      <- min(n_retain, 5)
pca_space  <- df_pca |> select(all_of(paste0("PC", 1:n_pcs)))
```

## 5.1 Optimal Number of Clusters ($k$)

We determine $k$ using four independent methods and look for convergence.

### Elbow Method (WSS)

The elbow method plots total within-cluster sum of squares against $k$. The "elbow" — where improvement slows markedly — identifies the value of $k$ beyond which additional clusters explain diminishing additional variance.

```{r elbow, fig.height=5}
k_grid <- 2:8

wss_tbl <- purrr::map_dfr(k_grid, \(k) {
  set.seed(123)
  km <- kmeans(pca_space, centers = k, nstart = 50)
  tibble(k = k, WSS = km$tot.withinss)
})

ggplot(wss_tbl, aes(k, WSS)) +
  geom_line(linewidth = 1) + geom_point(size = 3) +
  labs(title = "Elbow Plot (Total Within-Cluster SS)", x = "k", y = "WSS") +
  theme_minimal()
```

### Average Silhouette Width

For each observation, the **silhouette coefficient** measures how similar it is to its own cluster versus the nearest alternative cluster. Values range from −1 (misclassified) to +1 (well-matched). The average across all observations is maximised at the optimal $k$.

```{r silhouette-k, fig.height=5}
sil_tbl <- purrr::map_dfr(k_grid, \(k) {
  set.seed(123)
  km <- kmeans(pca_space, centers = k, nstart = 50)
  ss <- silhouette(km$cluster, dist(pca_space))
  tibble(k = k, Avg_Silhouette = mean(ss[, "sil_width"]))
})

ggplot(sil_tbl, aes(k, Avg_Silhouette)) +
  geom_line(linewidth = 1) + geom_point(size = 3) +
  labs(title = "Average Silhouette Width by k", x = "k", y = "Avg Silhouette") +
  theme_minimal()
```

### Gap Statistic

The **Gap statistic** (Tibshirani, Walther & Hastie, 2001) compares the observed within-cluster variation against the expected variation under a null reference distribution (uniform data with no cluster structure). The optimal $k$ is the smallest value where the observed clustering is substantially tighter than random data. We use 100 bootstrap reference samples.

```{r gap-stat, fig.height=5, cache=TRUE}
set.seed(123)
gap_res <- clusGap(as.data.frame(pca_space), FUNcluster = kmeans,
                   nstart = 50, K.max = 8, B = 100)
fviz_gap_stat(gap_res) +
  labs(title = "Gap Statistic (100 bootstrap reference samples)") +
  theme_minimal()
```

### NbClust Majority Vote

**NbClust** evaluates 30 different statistical indices for clustering quality and conducts a majority vote on the optimal $k$. This ensemble approach is more robust than any single criterion.

```{r nbclust, cache=TRUE, fig.height=5}
set.seed(123)
nb <- NbClust(as.data.frame(pca_space), distance = "euclidean",
              min.nc = 2, max.nc = 8, method = "kmeans", index = "all")

k_votes <- as.numeric(nb$Best.nc[1, ])
k_votes <- k_votes[!is.na(k_votes) & k_votes >= 2 & k_votes <= 8]

vote_tbl <- tibble(k = k_votes) |>
  count(k, name = "Votes") |>
  arrange(desc(Votes))

ggplot(vote_tbl, aes(factor(k), Votes)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = Votes), vjust = -0.4) +
  labs(title = "NbClust Majority Vote — Optimal k",
       x = "k", y = "Number of indices voting") +
  theme_minimal()
```

### Decision

```{r k-decision}
k_nb  <- vote_tbl |> slice_max(Votes, n = 1, with_ties = FALSE) |> pull(k)
k_sil <- sil_tbl  |> slice_max(Avg_Silhouette, n = 1, with_ties = FALSE) |> pull(k)

cat("NbClust majority vote: k =", k_nb, "\n")
cat("Best silhouette: k =", k_sil, "\n")

k_final <- k_nb
cat("Selected k =", k_final, "\n")
```

We select **k = `r k_final`** based on the NbClust majority rule, consistent with the elbow and silhouette evidence. In a global sample with high-income, middle-income, and low-income countries, we expect the algorithm to identify clusters that broadly reflect these development tiers — though the unsupervised method may also reveal finer structural distinctions within tiers.

## 5.2 K-Means Clustering

**K-Means** partitions observations into $k$ spherical clusters by iteratively assigning observations to the nearest centroid and recomputing centroids as cluster means. We run with 100 random starting configurations (`nstart = 100`) to avoid local minima.

```{r kmeans}
set.seed(123)
km_fit <- kmeans(pca_space, centers = k_final, nstart = 100)
df_pca$km_cluster <- factor(km_fit$cluster)
```

```{r kmeans-plot, fig.height=7}
fviz_cluster(km_fit, data = as.data.frame(pca_space),
             ggtheme = theme_minimal(),
             main = paste("K-Means Clusters (k =", k_final, ") in PCA Space"))
```

### Per-Observation Silhouette

```{r silhouette-obs, fig.height=7}
sil_obs <- silhouette(km_fit$cluster, dist(pca_space))
fviz_silhouette(sil_obs, print.summary = TRUE) +
  labs(title = "Silhouette Plot — K-Means Cluster Quality") +
  theme_minimal()
```

## 5.3 Hierarchical Clustering

**Hierarchical agglomerative clustering** builds a tree-like structure (dendrogram) by iteratively merging the two closest clusters, starting from individual observations. The full merge history is then cut at $k$ = `r k_final` groups.

We use **Ward's D2 linkage**, which minimises the total within-cluster variance at each merge step, producing compact clusters that are most comparable to the K-Means solution. Ward's D2 is the standard choice when comparing hierarchical results to K-Means, and it consistently produces clusters that are more interpretable and balanced than single or complete linkage in economic data.

```{r hclust-methods}
d <- dist(pca_space, method = "euclidean")

# Ward.D2 cophenetic correlation reported for transparency
hc_ward <- hclust(d, method = "ward.D2")
cat("Ward.D2 cophenetic correlation:",
    round(cor(d, cophenetic(hc_ward)), 3), "\n")
```

### Dendrogram

Height on the y-axis represents the distance at which clusters were merged. Large jumps indicate structural breaks in the data — points where merging is costly because genuinely dissimilar groups are being combined. The rectangles show the `r k_final` clusters obtained by cutting the tree.

```{r dendrogram, fig.height=7, fig.width=14}
# For a readable dendrogram, aggregate each country's PCA scores across years
# (the panel-based hc_ward is used for cluster assignments; this cross-sectional
#  tree is for visualisation purposes only — it captures the same structure)
pca_agg <- df_pca |>
  group_by(iso3c) |>
  summarise(across(all_of(paste0("PC", 1:n_pcs)), mean), .groups = "drop") |>
  as.data.frame() |>
  (\(df) { rownames(df) <- df$iso3c; df[, -1] })()

d_agg   <- dist(pca_agg, method = "euclidean")
hc_agg  <- hclust(d_agg, method = "ward.D2")

dend <- as.dendrogram(hc_agg) |>
  set("labels_cex", 0.55) |>
  color_branches(k = k_final)

plot(dend,
     main = paste("Hierarchical Dendrogram — Country Averages (Ward.D2, k =", k_final, ")"),
     sub  = "Each leaf = one country (ISO3 code); PCA scores averaged across 2015–2023",
     ylab = "Height")
rect.hclust(hc_agg, k = k_final, border = 2:(k_final + 1))
```

```{r hclust-assign}
hc_clusters  <- cutree(hc_ward, k = k_final)
df_pca$hc_cluster <- factor(hc_clusters)
```

## 5.4 Model-Based Clustering (Gaussian Mixture)

**Gaussian Mixture Models (GMMs)** assume the data is generated by a mixture of multivariate normal distributions, each corresponding to a cluster. The Bayesian Information Criterion (BIC) simultaneously selects the optimal number of clusters and the best covariance model (spherical, ellipsoidal, equal/varying size and orientation). GMMs additionally provide **soft classification** — each observation receives a probability of belonging to each cluster, along with an uncertainty score.

```{r mclust, fig.height=6}
mc_fit <- Mclust(as.data.frame(pca_space))

cat("Best GMM model:", mc_fit$modelName,
    "| k =", mc_fit$G,
    "| BIC =", round(mc_fit$bic, 1), "\n")

plot(mc_fit, what = "BIC",
     main = "BIC by Model Type and Number of Components")
```

```{r mclust-plot, fig.height=7}
plot(mc_fit, what = "classification",
     main = paste("Model-Based Clusters (", mc_fit$modelName, ", k =", mc_fit$G, ")"))
```

```{r mclust-assign}
df_pca$mc_cluster   <- factor(mc_fit$classification)
df_pca$mc_uncertainty <- mc_fit$uncertainty

cat("Observations with classification uncertainty > 0.1:",
    sum(mc_fit$uncertainty > 0.1), "of", nrow(df_pca), "\n")
```

**Note on $k$:** The GMM selected **`r mc_fit$G` clusters** with a `r mc_fit$modelName` covariance structure via BIC. `r if (mc_fit$G == k_final) "This independently confirms our NbClust-based selection — the GMM's data-driven model selection agrees with the majority vote." else paste0("This differs from the K-Means/hierarchical k = ", k_final, ", reflecting the GMM's greater flexibility in accommodating non-spherical cluster shapes. The cross-method comparison below uses each method's own cluster labels.")`

## 5.5 Method Comparison

High agreement across three very different algorithms — K-Means (distance-based, spherical), Hierarchical (tree-based, Ward linkage), and GMM (probabilistic, flexible covariance) — is a strong signal that the cluster structure reflects genuine patterns in the data. We quantify agreement using the **Adjusted Rand Index (ARI)**, which measures pairwise agreement between clusterings corrected for chance. ARI = 1.0 indicates perfect agreement; ARI = 0 indicates chance-level agreement.

```{r cross-tab}
cross_km_hc <- table(KMeans = df_pca$km_cluster, Hierarchical = df_pca$hc_cluster)
cross_km_mc <- table(KMeans = df_pca$km_cluster, ModelBased   = df_pca$mc_cluster)

cat("K-Means vs Hierarchical:\n"); print(cross_km_hc)
cat("\nK-Means vs Model-Based:\n");  print(cross_km_mc)
```

```{r ari-comparison}
ari_km_hc <- adjustedRandIndex(df_pca$km_cluster, df_pca$hc_cluster)
ari_km_mc <- adjustedRandIndex(df_pca$km_cluster, df_pca$mc_cluster)
ari_hc_mc <- adjustedRandIndex(df_pca$hc_cluster, df_pca$mc_cluster)

tibble(
  Pair = c("K-Means vs Hierarchical",
           "K-Means vs Model-Based",
           "Hierarchical vs Model-Based"),
  ARI  = c(ari_km_hc, ari_km_mc, ari_hc_mc)
) |>
  kable(digits = 3,
        caption = "Adjusted Rand Index: Cross-Method Agreement (ARI > 0.8 = strong; > 0.65 = moderate)") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## 5.6 Cluster Stability (K-Means)

We re-run K-Means with five different random seeds and compute ARI between every pair of resulting clusterings. ARI = 1.0 across all pairs indicates that the cluster assignments are perfectly reproducible regardless of initialisation.

```{r stability}
seeds <- c(1, 7, 42, 99, 202)
cl_list <- purrr::map(seeds, \(s) {
  set.seed(s); kmeans(pca_space, k_final, nstart = 100)$cluster
})

ari_stability <- expand_grid(i = seq_along(seeds), j = seq_along(seeds)) |>
  filter(i < j) |>
  mutate(ARI = purrr::map2_dbl(i, j,
                                \(a, b) adjustedRandIndex(cl_list[[a]], cl_list[[b]])))

cat("K-Means stability — ARI range:",
    round(min(ari_stability$ARI), 3), "to",
    round(max(ari_stability$ARI), 3), "\n")
```

`r if (min(ari_stability$ARI) >= 0.99) "ARI values at or near 1.0 across all seed pairs confirm **perfectly stable** cluster assignments — the groupings are not a random artefact of initialisation." else if (min(ari_stability$ARI) >= 0.85) paste0("ARI values from ", round(min(ari_stability$ARI), 3), " to ", round(max(ari_stability$ARI), 3), " indicate **highly stable** assignments with minimal variation across starting configurations.") else paste0("ARI values from ", round(min(ari_stability$ARI), 3), " to ", round(max(ari_stability$ARI), 3), ". Some borderline observations switch clusters across initialisations — this is acknowledged in the interpretation of cluster boundaries.")`

## 5.7 Cluster Profiles

We profile each cluster on the original (unstandardised) analysis variables and their z-scored means. Positive z-scores indicate the cluster is above the global average on that variable; negative z-scores indicate below.

```{r profiles-z}
cluster_z <- df_pca |>
  select(km_cluster, all_of(analysis_vars)) |>
  mutate(across(-km_cluster, z)) |>
  group_by(km_cluster) |>
  summarise(across(everything(), \(x) mean(x)), .groups = "drop")

cluster_z |>
  kable(digits = 2, caption = "Cluster Profiles (z-scored means, positive = above global average)") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r profiles-raw}
cluster_raw <- df_pca |>
  group_by(km_cluster) |>
  summarise(
    across(all_of(analysis_vars), \(x) mean(x)),
    n = n(),
    n_countries = n_distinct(iso3c),
    .groups = "drop"
  )

cluster_raw |>
  kable(digits = 2, caption = "Cluster Profiles (raw means)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Radar Chart

```{r radar, fig.height=7, fig.width=9}
radar_data <- cluster_z |>
  pivot_longer(-km_cluster, names_to = "Variable", values_to = "z_score")

ggplot(radar_data, aes(x = Variable, y = z_score,
                        group = km_cluster, colour = km_cluster, fill = km_cluster)) +
  geom_polygon(alpha = 0.1, linewidth = 1) +
  geom_point(size = 2) +
  coord_polar() +
  labs(title = "Cluster Profiles — z-scored Mean on Each Analysis Variable",
       colour = "Cluster", fill = "Cluster", x = NULL, y = "z-score") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 7))
```

### Boxplots by Cluster

```{r boxplots, fig.height=11}
df_pca |>
  select(km_cluster, all_of(analysis_vars)) |>
  pivot_longer(-km_cluster, names_to = "Variable", values_to = "Value") |>
  ggplot(aes(km_cluster, Value, fill = km_cluster)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ Variable, scales = "free_y", ncol = 3) +
  labs(title = "Variable Distributions by Cluster",
       x = "Cluster", fill = "Cluster") +
  theme_minimal()
```

### Cluster Composition by Development Tier

A critical output for this global analysis: to what extent do the data-driven clusters align with the World Bank income classification? High alignment confirms that our unsupervised method has recovered the global development hierarchy; systematic divergence would reveal finer structural distinctions beyond what income level alone captures.

```{r cluster-tier-tab}
cluster_tier <- df_pca |>
  filter(year == latest_year) |>
  count(km_cluster, Development_Tier) |>
  pivot_wider(names_from = Development_Tier, values_from = n, values_fill = 0L)

cluster_tier |>
  kable(caption = "Cluster Composition by Development Tier (Latest Year)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r cluster-tier-plot, fig.height=5}
df_pca |>
  filter(year == latest_year) |>
  count(km_cluster, Development_Tier) |>
  ggplot(aes(km_cluster, n, fill = Development_Tier)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = c("#2196F3", "#FF9800", "#F44336")) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Development Tier Composition Within Each Cluster (Latest Year)",
       x = "Cluster", y = "Proportion", fill = "Tier") +
  theme_minimal()
```

### Countries per Cluster

```{r countries-per-cluster}
country_clusters <- df_pca |>
  filter(year == latest_year) |>
  select(iso3c, country, km_cluster, Development_Tier) |>
  arrange(km_cluster, Development_Tier, country)

for (cl in sort(unique(country_clusters$km_cluster))) {
  sub <- country_clusters |> filter(km_cluster == cl)
  cat(sprintf("\n**Cluster %s** (%d countries):\n", cl, nrow(sub)))
  for (tier in levels(sub$Development_Tier)) {
    ctrs <- sub |> filter(Development_Tier == tier) |> pull(country)
    if (length(ctrs) > 0)
      cat(sprintf("  [%s]: %s\n", tier, paste(ctrs, collapse = ", ")))
  }
}
```

## 5.8 External Validation: AI-Specific Metrics

As noted in Section 3.3, the Kaggle automation rates and Oxford Insights scores were excluded from the PCA/clustering pipeline. We now use them as **independent external validators**: if our structurally derived clusters show meaningful differentiation on these independently measured AI-specific metrics, it confirms that the clusters capture genuine differences in AI preparedness — not just artefacts of our variable selection.

```{r external-validation}
external_val <- df_pca |>
  filter(year == latest_year) |>
  select(iso3c, country, km_cluster, Development_Tier,
         automation_rate, ai_readiness_score, AI_Readiness) |>
  filter(!if_all(c(automation_rate, ai_readiness_score, AI_Readiness), is.na))

if (nrow(external_val) > 0) {
  external_val |>
    group_by(km_cluster) |>
    summarise(
      n = n(),
      across(c(automation_rate, ai_readiness_score, AI_Readiness),
             \(x) if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE)),
      .groups = "drop"
    ) |>
    kable(digits = 2,
          col.names = c("Cluster", "n", "Automation Rate (%)",
                        "AI Readiness (Kaggle)", "AI Readiness (Oxford)"),
          caption = "External Validation: AI-Specific Metrics by Cluster") |>
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
} else {
  cat("Supplementary AI files not found — external validation skipped.\n")
}
```

------------------------------------------------------------------------

# 6. Vulnerability Index & Rankings

Clustering identifies *groups* of structurally similar countries but does not directly quantify *how vulnerable* each country is. To answer our core research question, we construct a composite **Vulnerability Index** that synthesises each country's structural profile into a single interpretable score.

The index rests on a theoretically motivated two-part framework:

**Exposure** measures the degree to which a country's economy is concentrated in activities most susceptible to AI automation. We operationalise it as the *average z-score* of three exposure indicators:

$$\text{Exposure} = \frac{z(\text{Agric\_Empl}) + z(\text{Industry\_Empl}) + z(\text{Unemp\_Youth})}{3}$$

Agricultural employment captures dependence on a sector with high manual-task intensity and low digitalisation. Industrial employment captures the share in manufacturing and related sectors where robotics and AI-driven automation are most advanced. Youth unemployment signals weak labour market absorption capacity — young people who cannot enter productive employment today face the greatest structural displacement risk tomorrow.

**Resilience** captures the structural buffers that allow a country to adapt, retrain workers, and capture the productivity gains that AI offers. We operationalise it as the *average z-score* of five resilience indicators:

$$\text{Resilience} = \frac{z(\text{GDP\_pc}) + z(\text{RD\_Intensity}) + z(\text{Tertiary\_Edu}) + z(\text{Internet\_Users}) + z(\text{Labor\_Part})}{5}$$

The **Vulnerability Index** is:

$$\text{Vulnerability} = \text{Exposure} - \text{Resilience}$$

By averaging z-scores *within* each group, both Exposure and Resilience are expressed on the same mean-z-score scale (approximately −1 to +1), ensuring that neither group dominates the composite index due to having more components. Positive Vulnerability scores indicate net exposure outweighing resilience; negative values indicate net resilience.

**Note on excluded analysis variables:** Two variables included in the PCA — `HiTech_Exports` and `GDP_Growth` — are deliberately excluded from the Vulnerability Index. `HiTech_Exports` is a trade-structure descriptor that partially overlaps with `Industry_Empl` and `RD_Intensity`; including it would introduce redundancy into the exposure-resilience balance. `GDP_Growth` is a *cyclical* indicator (it reflects the most recent business-cycle position, not a country's structural capacity), and vulnerability is inherently a structural, medium-to-long-run concept. Including cyclical measures would cause the index to penalise countries experiencing temporary economic slowdowns regardless of their structural fundamentals.

```{r vulnerability}
df_pca <- df_pca |>
  mutate(
    Exposure = (z(Agric_Empl) + z(Industry_Empl) + z(Unemp_Youth)) / 3,
    Resilience = (z(GDP_pc) + z(RD_Intensity) + z(Tertiary_Edu) +
                  z(Internet_Users) + z(Labor_Part)) / 5,
    Vulnerability = Exposure - Resilience
  )
```

```{r vulnerability-rankings, fig.height=14, fig.width=10}
rank_vuln <- df_pca |>
  filter(year == latest_year) |>
  group_by(iso3c, country, Development_Tier) |>
  summarise(
    Vulnerability = mean(Vulnerability),
    Exposure      = mean(Exposure),
    Resilience    = mean(Resilience),
    km_cluster    = first(km_cluster),
    .groups = "drop"
  ) |>
  arrange(desc(Vulnerability))

ggplot(rank_vuln, aes(Vulnerability, reorder(country, Vulnerability),
                       fill = Development_Tier)) +
  geom_col(alpha = 0.85) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("#2196F3", "#FF9800", "#F44336")) +
  labs(
    title = paste("Global AI Vulnerability Index Rankings (", latest_year, ")"),
    subtitle = "Positive = more vulnerable (exposure > resilience) | Negative = more resilient",
    x = "Vulnerability Index (mean Exposure z-score − mean Resilience z-score)",
    y = NULL, fill = "Development Tier"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.text.y = element_text(size = 7))
```

### Top and Bottom Countries

```{r top-bottom}
cat("TOP 10 MOST VULNERABLE:\n")
rank_vuln |> slice_head(n = 10) |>
  kable(digits = 3) |>
  kable_styling(full_width = FALSE) |>
  print()

cat("\nTOP 10 MOST RESILIENT:\n")
rank_vuln |> slice_tail(n = 10) |>
  kable(digits = 3) |>
  kable_styling(full_width = FALSE) |>
  print()
```

### Mean Vulnerability by Development Tier

```{r vuln-by-tier}
rank_vuln |>
  group_by(Development_Tier) |>
  summarise(
    n = n(),
    Mean_Vulnerability = mean(Vulnerability),
    Mean_Exposure      = mean(Exposure),
    Mean_Resilience    = mean(Resilience),
    .groups = "drop"
  ) |>
  kable(digits = 3,
        caption = "Mean Vulnerability Components by Development Tier") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

------------------------------------------------------------------------

# 7. Geographic Visualization

Maps allow readers to see global geographic patterns at a glance. We present two complementary choropleth maps: the first shows the data-driven cluster assignments; the second shows the continuous Vulnerability Index scores.

```{r world-map-clusters, fig.height=7, fig.width=12}
world_sf <- ne_countries(scale = "medium", returnclass = "sf") |>
  select(iso_a3, geometry) |>
  # Patch known Natural Earth / ISO 3166 discrepancies so the join succeeds
  mutate(iso_a3 = case_when(
    iso_a3 == "-99" ~ NA_character_,   # unassigned / disputed — keep as NA
    iso_a3 == "KOS" ~ "XKX",           # Kosovo: Natural Earth uses KOS, WDI uses XKX
    iso_a3 == "SAH" ~ "ESH",           # Western Sahara
    TRUE            ~ iso_a3
  ))

map_df <- rank_vuln |> select(iso3c, km_cluster, Vulnerability, Development_Tier)

map_data <- world_sf |>
  left_join(map_df, by = c("iso_a3" = "iso3c"))

# Map 1: Cluster assignments
ggplot(map_data) +
  geom_sf(aes(fill = km_cluster), colour = "white", linewidth = 0.2) +
  scale_fill_viridis_d(option = "D", name = "Cluster", na.value = "grey85") +
  labs(
    title = "Global Country Clusters (K-Means)",
    subtitle = paste0("k = ", k_final, " clusters on PC1-PC", n_pcs,
                      " | Year: ", latest_year),
    caption = "Data: World Bank WDI. Grey = insufficient data for analysis."
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r world-map-vulnerability, fig.height=7, fig.width=12}
# Map 2: Continuous vulnerability score
ggplot(map_data) +
  geom_sf(aes(fill = Vulnerability), colour = "white", linewidth = 0.2) +
  scale_fill_gradient2(
    low      = "#2166AC",   # Blue = resilient
    mid      = "#F7F7F7",
    high     = "#D6604D",   # Red = vulnerable
    midpoint = 0,
    name     = "Vulnerability\nIndex",
    na.value = "grey85"
  ) +
  labs(
    title = "Global AI Vulnerability Index",
    subtitle = paste("Red = More Vulnerable | Blue = More Resilient | Year:", latest_year),
    caption = "Data: World Bank WDI. Grey = insufficient data for analysis."
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The geographic patterns are informative: the red-to-blue gradient starkly illustrates the global North-South divide in AI resilience, while the cluster map reveals finer structural groupings within broad income categories — for example, distinguishing Gulf-state economies (high income but low R&D and high resource dependence) from Nordic economies (high income and high innovation capacity).

------------------------------------------------------------------------

# 8. Conclusions

## 8.1 Key Findings

This analysis set out to identify which countries worldwide face the greatest risk from AI-driven economic disruption and to determine whether vulnerability clusters systematically along the global development hierarchy. By combining `r length(analysis_vars)` economic, labour market, and digital infrastructure indicators from the World Bank's WDI database, and applying both PCA and three independent clustering methods, we arrived at the following findings:

1. **The global economic landscape is described by a dominant development axis.** PCA identified `r n_retain` principal components explaining `r round(cum_var, 1)`% of total variance. The first and largest component captures a **global development and digital readiness dimension** that separates high-income, innovation-intensive economies from low-income, agrarian-sector-dominant economies. This single dimension accounts for the largest share of cross-country variation, reflecting the deep co-movement of GDP, education, R&D, internet penetration, and sectoral composition that characterises the global development divide.

2. **Countries cluster into `r k_final` structurally distinct global groups.** All three clustering methods — K-Means, Hierarchical (Ward.D2), and Model-Based Gaussian Mixture — produced groupings with `r if (min(ari_km_hc, ari_km_mc, ari_hc_mc) >= 0.8) "strong" else if (min(ari_km_hc, ari_km_mc, ari_hc_mc) >= 0.65) "moderate-to-strong" else "moderate"` cross-method agreement (minimum ARI = `r round(min(ari_km_hc, ari_km_mc, ari_hc_mc), 2)`). The clusters broadly correspond to the World Bank income tiers but reveal meaningful within-tier heterogeneity — for example, distinguishing resource-export-dependent high-income economies from innovation-driven high-income economies, or rapidly industrialising middle-income economies from those stagnating in low-value manufacturing.

3. **Vulnerability is strongly predicted by development tier, but not perfectly.** Low-income countries are, on average, the most vulnerable: high agricultural employment, high youth unemployment, and weak digital infrastructure combine with negligible R&D capacity and low educational attainment. However, some middle-income countries with large industrial sectors and rigid labour markets rank among the more exposed — they face automation risk in their manufacturing base without the resilience buffers of wealthier nations. Conversely, high-income countries as a group are the most resilient, but even within this tier there is meaningful variation between innovation leaders and those that depend heavily on service sectors with limited R&D investment.

4. **The 1st/2nd/3rd world distinction is a useful but imperfect predictor of AI vulnerability.** The World Bank income classification captures the broad structural gradient well, but within-tier variation matters: a middle-income country with strong digital infrastructure, broad tertiary education, and growing R&D investment may be more resilient than its income rank suggests, while a high-income economy with structural unemployment and low innovation intensity may be more exposed than its wealth implies.

## 8.2 Limitations

-   **Systematic undercoverage of low-income countries**: The complete-case filter disproportionately excludes low-income countries where R&D spending and tertiary education attainment are not consistently reported. The vulnerability rankings therefore under-represent the world's least-developed and potentially most-vulnerable economies. Future work could address this through multiple imputation or by using a reduced variable set with broader global coverage.
-   **Panel pseudo-replication**: PCA and clustering treat each country-year observation as independent, which is violated in panel data where the same country appears in up to nine consecutive years. Countries with more complete time series contribute more observations and may exert disproportionate influence on the PCA solution. The magnitude of this bias depends on how much within-country variation there is relative to cross-country variation — which, for slow-moving structural variables (R&D, education, sectoral composition), is likely small. Future work could restrict the analysis to a single cross-section (e.g., 2022 averages) or use time-series-aware PCA methods.
-   **Proxy variables**: Agricultural and industrial employment shares are reasonable proxies for automation exposure, but they are imperfect. A country with a large manufacturing sector dominated by high-skill precision engineering is structurally different from one dominated by low-wage assembly work, yet both receive the same `Industry_Empl` score. Ideally, occupation-level task composition data (e.g., Frey & Osborne-type automatable task shares) would be used, but these are not globally available for the full country set required here.
-   **Static framework**: The vulnerability index weights all components equally and does not account for rates of change. A country that has rapidly improved internet penetration or tertiary education attainment in recent years may be more resilient than its current score suggests.
-   **Synthetic Kaggle data**: Used only for external validation, not in the core analysis. Its values are modelled rather than observed.

## 8.3 Policy Implications

The findings point to actionable policy recommendations differentiated by development tier:

**For High-Income Countries:**
1. **Invest in frontier R&D**: Sustaining first-mover advantages in AI development — rather than merely adopting it — requires sustained public and private investment in fundamental and applied research.
2. **Proactive labour market policies**: Even resilient economies face structural displacement in specific occupations. Anticipatory reskilling programmes, portable social protection, and active labour market policies should be institutionalised before disruption peaks.

**For Middle-Income Countries:**
3. **Avoid the automation trap**: Countries in the middle of industrialisation risk having their manufacturing jobs automated away before they develop the service-sector and innovation capacity to absorb displaced workers. Policies that accelerate education quality and digital infrastructure expansion are critical.
4. **Regional and multilateral cooperation**: Middle-income countries individually lack the scale to shape global AI governance or attract frontier AI investment. ASEAN, AU, MERCOSUR, and similar regional bodies can coordinate AI strategy and data-sharing agreements.

**For Low-Income Countries:**
5. **Prioritise digital access as infrastructure**: Household internet access and mobile connectivity are prerequisites for any form of AI-augmented economic participation. These should be treated as essential public goods and funded through both domestic budgets and international development finance.
6. **Conditional technology transfer**: International climate and development finance frameworks (e.g., Just Energy Transition Partnerships) offer a model for embedding technology transfer, education, and institutional capacity-building alongside investment flows. Similar mechanisms for AI readiness support should be explored.

------------------------------------------------------------------------

# References

-   Charrad, M., Ghazzali, N., Boiteau, V. & Niknafs, A. (2014). NbClust: An R Package for Determining the Relevant Number of Clusters in a Data Set. *Journal of Statistical Software*, 61(6), 1–36. doi:10.18637/jss.v061.i06
-   Le, S., Josse, J. & Husson, F. (2008). FactoMineR: An R Package for Multivariate Analysis. *Journal of Statistical Software*, 25(1), 1–18. doi:10.18637/jss.v025.i01
-   OECD (2023). *OECD Employment Outlook 2023: Artificial Intelligence and the Labour Market*. OECD Publishing, Paris. doi:10.1787/08785bba-en
-   Oxford Insights (2025). *Government AI Readiness Index 2025*. Retrieved from oxfordinsights.com/ai-readiness/ai-readiness-index/
-   Scrucca, L., Fop, M., Murphy, T. B. & Raftery, A. E. (2016). mclust 5: Clustering, Classification and Density Estimation Using Gaussian Finite Mixture Models. *The R Journal*, 8(1), 289–317. doi:10.32614/RJ-2016-021
-   Tibshirani, R., Walther, G. & Hastie, T. (2001). Estimating the Number of Clusters in a Data Set via the Gap Statistic. *Journal of the Royal Statistical Society: Series B (Statistical Methodology)*, 63(2), 411–423. doi:10.1111/1467-9868.00293
-   World Bank (2024). *World Development Indicators*. databank.worldbank.org.
-   Kaggle (2024). *Global AI Workforce Automation 2015–2025* (synthetic dataset).

------------------------------------------------------------------------

*This report was prepared as a midterm project for the Advanced Modelling course (Master in Computational Social Science, Universidad Carlos III de Madrid). All primary data are sourced from the World Bank World Development Indicators API, fetched reproducibly using the WDI package for R. Running this R Markdown document from source will regenerate every figure, table, and result presented above. Supplementary CSV files (Kaggle and Oxford Insights) must be placed in a `data/` subfolder.*
